---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sporthelp

The sporthelp package aims to help users in doing calculations based around
fitness. It provides functions to do common calculations without having to
write them out in scripts.

## Installation

You can install the development version of sporthelp from
[GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("stevenballin/sporthelp")
```

## Functions

```{r, echo = FALSE}
knitr::kable(data.frame(
  Function = c(
    "`calculate_bmi()`",
    "`calculate_bmi_prime()`",
    "`calculate_ponderal_index()`",
    "`daily_calories()`"
  ),
  Description = c(
    "Calculate Body Mass Index (BMI)",
    "Calculate BMI Prime relative to a continental or custom upper limit",
    "Calculate Ponderal Index (PI)",
    "Calculate recommended daily calorie intake"
  )
))
```

## Usage

```{r}
library(sporthelp)
```

### BMI

Body Mass Index is calculated from weight and height. Rather than writing the
formula manually (error prone), use `calculate_bmi()`:

```{r}
# Manual approach
weight_kg <- 80
height_cm <- 180
weight_kg / (height_cm / 100)^2

# Using sporthelp
calculate_bmi(weight = 80, height = 180)
```

Height and weight units can be specified explicitly:

```{r}
calculate_bmi(weight = 176, height = 71, weight_unit = "lbs", height_unit = "in")
```

---

### BMI Prime

BMI Prime expresses BMI as a ratio relative to the upper limit for a given
region, where a value of 1.0 means BMI is exactly at the upper limit.
It can be calculated from a known BMI or directly from height and weight:

```{r}
# From a known BMI
calculate_bmi_prime(bmi = 25, continent = "Europe")

# From height and weight
calculate_bmi_prime(weight = 80, height = 180, continent = "Asia")

# With a custom upper limit
calculate_bmi_prime(bmi = 25, upper_limit = 25)
```

Supported continents:

```{r, echo = FALSE}
knitr::kable(data.frame(
  Continent = c("Asia", "Europe", "North_America", "South_America", "Africa", "Oceania"),
  `Upper Limit` = c(22.9, 24.9, 24.9, 24.9, 24.9, 25.0),
  check.names = FALSE
))
```

---

### Ponderal Index

The Ponderal Index is similar to BMI but cubes height instead of squaring it,
making it more reliable for very tall or short individuals:

```{r}
# Default units (kg, cm)
calculate_ponderal_index(weight = 80, height = 180)

# Imperial units
calculate_ponderal_index(weight = 160, height = 70, weight_unit = "lbs", height_unit = "in")
```

---

### Daily Calories

Calculates recommended daily calorie intake based on weight, height, age,
gender and activity level. Three BMR formulas are supported:

- `"MSJ"` — Mifflin-St Jeor (default)
- `"RHB"` — Revised Harris-Benedict
- `"KMA"` — Katch-McArdle (requires body fat percentage as a decimal e.g. `0.20` for 20%)

```{r}
# Default formula (MSJ) with Light activity
daily_calories(weight = 80, height = 180, age = 30, gender = "male")

# Specifying activity level
daily_calories(weight = 60, height = 165, age = 25, gender = "female", activity = "Active")

# Using the Katch-McArdle formula
daily_calories(weight = 80, height = 180, age = 30, gender = "male",
               bodyfat = 0.20, formula = "KMA")

# Imperial units
daily_calories(weight = 176, height = 71, age = 30, gender = "male",
               weight_unit = "lbs", height_unit = "in")
```

Supported activity levels:

```{r, echo = FALSE}
knitr::kable(data.frame(
  Level = c("BMR", "Sedentary", "Light", "Moderate", "Active", "Very Active", "Extra Active"),
  Factor = c(1.000, 1.200, 1.375, 1.465, 1.550, 1.725, 1.900)
))
```
