---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# sporthelp

The sporthelp package aims to help users in doing calculations based around fitness. 
It provides functions to do common calculations without having to write them out in scripts.

## Installation

You can install the development version of sporthelp from [GitHub](https://github.com/) with:
      
``` r
# install.packages("devtools")
devtools::install_github("stevenballin/sporthelp")
```

## Usage

When doing calculations based around fitness you may want to do calculations surrounding BMI, daily calories or step goals. Rather than typing these calculations out as below (error prone) the alternative solution is to use the functions supplied in this package.
```{r}
weight_kg <- 80
height_cm <- 180
weight_kg/(height_cm/100)^2
```

Now below one of the base functions of sporthelp is shown:

```{r}
library(sporthelp)
calculate_bmi(weight=80, height=180)
```
