---
title: "Getting Started with sporthelp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sporthelp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sporthelp)
```

## Introduction

The sporthelp package provides functions for common fitness calculations.
Rather than writing error-prone formulas by hand, sporthelp handles unit
conversions, input validation, and the underlying maths for you.

This vignette walks through each function with realistic examples, and shows
how functions can be combined to build a fuller picture of a person's fitness
metrics.

---

## BMI

Body Mass Index is the most widely used measure of leanness or corpulence.
It is calculated as weight in kilograms divided by height in metres squared.

```{r}
# A 30 year old male, 80kg, 180cm
calculate_bmi(weight = 80, height = 180)
```

Units can be specified if your data is not in metric:

```{r}
# Same person in imperial units
calculate_bmi(weight = 176, height = 71, weight_unit = "lbs", height_unit = "in")
```

### Limitations of BMI

BMI is a useful population-level measure but has well-documented limitations
for individuals — particularly those who are very tall, very short, or have
high muscle mass. The Ponderal Index and BMI Prime (covered below) address
some of these limitations.

---

## BMI Prime

BMI Prime expresses BMI as a ratio of a person's BMI to the upper limit
recommended for their region. A value of exactly 1.0 means the person is
right at their regional upper limit; below 1.0 is under, above 1.0 is over.

```{r}
# From a known BMI
calculate_bmi_prime(bmi = 25, continent = "Europe")

# Directly from height and weight
calculate_bmi_prime(weight = 80, height = 180, continent = "Asia")
```

If your use case requires a custom upper limit rather than a continental one,
you can supply it directly:

```{r}
calculate_bmi_prime(bmi = 25, upper_limit = 24.9)
```

### Continental upper limits

Different regions use different BMI upper limits due to population-level
differences in body composition:

```{r, echo = FALSE}
knitr::kable(data.frame(
  Continent    = c("Asia", "Europe", "North_America", "South_America", "Africa", "Oceania"),
  `Upper Limit` = c(22.9, 24.9, 24.9, 24.9, 24.9, 25.0),
  check.names  = FALSE
))
```

---

## Ponderal Index

The Ponderal Index (PI) is similar to BMI but cubes height rather than
squaring it. This makes it more reliable for individuals at the extremes
of the height spectrum, where BMI tends to over- or underestimate body fat.

```{r}
# Default metric units
calculate_ponderal_index(weight = 80, height = 180)

# Imperial units
calculate_ponderal_index(weight = 160, height = 70,
                         weight_unit = "lbs", height_unit = "in")
```

---

## Daily Calories

`daily_calories()` estimates recommended daily calorie intake based on
Basal Metabolic Rate (BMR) scaled by an activity factor. Three BMR
formulas are available:

- **MSJ** (Mifflin-St Jeor) — the default, recommended for most people
- **RHB** (Revised Harris-Benedict) — an older but widely used formula
- **KMA** (Katch-McArdle) — accounts for body fat, useful for athletic individuals

```{r}
# Default settings: MSJ formula, Light activity
daily_calories(weight = 80, height = 180, age = 30, gender = "male")

# More active lifestyle
daily_calories(weight = 80, height = 180, age = 30, gender = "male",
               activity = "Very Active")

# Female example
daily_calories(weight = 60, height = 165, age = 25, gender = "female",
               activity = "Moderate")

# Using Katch-McArdle for an athletic individual
daily_calories(weight = 80, height = 180, age = 30, gender = "male",
               bodyfat = 0.12, formula = "KMA", activity = "Active")
```

### Activity levels

```{r, echo = FALSE}
knitr::kable(data.frame(
  Level   = c("BMR", "Sedentary", "Light", "Moderate", "Active", "Very Active", "Extra Active"),
  Factor  = c(1.000, 1.200, 1.375, 1.465, 1.550, 1.725, 1.900),
  Example = c(
    "Completely at rest",
    "Desk job, little or no exercise",
    "Light exercise 1-3 days per week",
    "Moderate exercise 4-5 days per week",
    "Daily exercise or intense exercise 3-4 times per week",
    "Intense exercise 6-7 times per week",
    "Very intense daily training or physical job"
  )
))
```

---

## Combining functions

Since all functions use the same unit arguments, they can be combined
naturally. Here is an example building a full picture for one individual:

```{r}
weight <- 80
height <- 180
age    <- 30
gender <- "male"

bmi      <- calculate_bmi(weight = weight, height = height)
bmi_p    <- calculate_bmi_prime(weight = weight, height = height, continent = "Europe")
pi       <- calculate_ponderal_index(weight = weight, height = height)
calories <- daily_calories(weight = weight, height = height,
                           age = age, gender = gender, activity = "Moderate")

cat(
  "BMI:             ", round(bmi, 2),      "\n",
  "BMI Prime:       ", round(bmi_p, 2),    "\n",
  "Ponderal Index:  ", round(pi, 2),       "\n",
  "Daily Calories:  ", round(calories),    "\n"
)
```
